% Lute fretting systems

\chapter{Lute Fretting Systems}

In the previous chapter, I set forth several different methods in which we can measure
ratios of intervals to determine the quality of their temperament.  In this chapter, I
will apply these methods to different fretting schemes and determine, from a
mathematical standpoint, what kind of tempered interval each fret represents.  The
overall picture for lute temperaments from the sixteenth through early seventeenth
centuries can be outlined as follows: Pythagorean tunings were used exclusively at the
beginning of the period, and retained some use for certain intervals later in the
period. By the mid-sixteenth century, lute players began to add different types of
tempered intervals into the existing Pythagorean scheme.  Some of these intervals were
meantone, some approximating equal temperament, and others were original in their
quality.  From the end of century onward, fretting systems remained mixed, or else
used methods of dividing the octave into twelve equal semitones; however, their
semitones did not always match those of modern-day equal temperament. The important
distinction between these systems and keyboard temperaments at the time was
that lute temperaments did not fall completely into one kind of tuning or temperament
and were specific to the instrument.

The majority of instructions published during this period provided the player
with a practical means of setting frets and did not dwell on theory.  Almost all of the
sources have very precise rules for determining fret placement that use measuring tools
such as the straight-edge and compass.  These were commonly used for making
geometrical divisions, and when used to calculate fret placement, could
divide a line into any number of equal parts.  The remainder of the sources are less
precise and instruct the player to place a fret somewhere near another until it
sounds agreeable, or to move a fret slightly in one direction or the other, without
providing exact measurements.

What follows is a comparison of several fretting systems that appeared in lute and
vihuela treatises. For those sources that have instructions on placing frets using a
compass, I have determined the exact ratio of the interval using the arithmetic methods 
outlined in the previous chapter.  The details of the calculations can be found in
the appendix.

\section{Pythagorean Tunings for Lute}

Starting in the 9th century, most if not all fretted instruments were tuned according to
Pythagorean ratios.  The placement of frets on these instruments yielded pure fifths, fourths, and octaves
as well as wholetones with a ratio of 9:8. \autocite[9]{ML:1} The result of this, as we
saw in the previous chapter, was that major thirds were much wider than pure.  What we
know of the lute prior to 1500 suggests that it was a four- or five-stringed instrument
tuned in fourths.  Such a tuning would accommodate a Pythagorean tuning quite well, but
after 1500, the lute's repertoire changed and a sixth string was added.  The new tuning of
the 6-course lute used a major third between the instrument's third and fourth courses
which exposes the major limitation of a Pythagorean system of tuning.

Because theory treatises in the early sixteenth century still advocated Pythagorean
systems of tuning because of its mathematical principles, lute players were left to
forge their own solutions to the tuning dilemma.  One of the mainstays of lute
repertoire throughout the sixteenth century was intabulated polyphonic music. The first
books of music printed by Petrucci in the early 1500s contain peices of this kind, and
composers of lute music would include many of
them in their publications. Because this music had many harmonies that relied on
thirds, a Pythagorean system of tuning would have presented some obvious shortcomings.
It is at this point that we see lute players begin a gradual shift away from
Pythagorean tunings and towards meantone temperaments.

During the first half of the sixteenth century, at least three different tuning methods
for lute were published.  Two of these kept with the existing Pythagorean tradition of
the day, while the third departed from that tradition and employed a kind of meantone
tuning similar to that which Pietro Aaron and Vicentino were developing at the. 
\textit{Epithoma musice instrumentalis}, published in 1530 by Oronce
Fin\'{e}, was one of the sources that retained Pythagorean principles.  Fin\'{e} was a
professor of mathematics at the University of Paris and mostly published theoretical
works about mathematics. It was his personal interest in music that compelled him to
write \textit{Epithoma}. Written in Latin, the book includes instructions for tuning a
lute, reading tablature and setting frets of a lute.  Pierre Attaingnant was a friend
of Fin\'{e}'s and it was probably this friendship that resulted in the treatise's
publication.  An additional set of lute instructions appear in Attaignant's publication
\textit{Tres breve et familiere introduction pour entendre et apprendre [...]} in 1529,
this time written in French.  Some stipulate that these were instructions from Fin\'{e}
and just translated, but this has never been fully proven.\autocite[1]{PV:1}

Another example of Pythagorean tuning appeared in a book published about the middle of
sixteenth-century.  The book was not about music but instead contained various topics
concerning France.  Appearing in 1556 and carrying the long title of \textit{Discours non
plus melancoliques que diverses, de choses mesmement, qui appartiennent a notre FRANCE: \&
a la fin La maniere de bien \& iustement entoucher les Lucs \& Guiternes}, it has been
attributed to Bonaventure des Periers, but some sources list it as an anonymous author.
The subject of tuning lutes and guitars appears in the last chapter of the book, which,
curiously, has nothing to do with any of the other chapters in the book. The final chapter
of \textit{Discours [...]} contains instructions for fretting and tuning these instruments
as well as a diagram of a mesolab, an ancient geometric tool that was first used to solve
the problem of dividing a string into equal semitones.

In his assessment of their fretting instructions, Mark Lindley has concluded that Both
Fin\'{e} and the anonymous \textit{Discours [...]}  preserved the same characteristics of
other Pythagorean tuning methods.  These included earlier theorists from the fifteenth
century such as Henri Arnaut, Johannes Legreze, Nicola Burzio and Franchino Gafurio, as
well as more contemporary theorists like Heinrich Schreiber of Erfurt and Pietro Aron
\autocite[11]{ML:1}. The features of these fretting schemes resulted in fourths and fifths
at ratios of 4:3 and 3:2 respectively, and wholetones with a ratio of 9:8.  Semitones
depended on which fret the measurements began.  In \textit{Discours [...]}, frets 2, 4 and
6 were a series of 9:8 ratios beginning with the nut, while frets 1, 3, 5 and 7 were
series of 9:8 ratios beginning with the seventh fret and moving backwards.  Fin\'{e}'s
method was slightly different, placing frets 6, 8 and 10 at 9:8 ratios starting from the
twelfth fret.

The inherent problem with these systems is that the fret placement is based purely on
mathematics and maintaining Pythagorean ratios without any deference to practical
musicianship. Fin\'{e} was not a musician, and Lindley suggests that that the author of
\textit{Discours [...]} was not either \autocite[11]{ML:1}.  Furthermore, after
attempting to play various pieces from the period, he concludes that ``it seems
doubtful to me that sensitive players would really have left the pythagorean scheme
unaltered.'' \autocite[13]{ML:1}  This becomes apparent in later sources from Juan
Bermudo and Sylvestro Ganassi who each began with a Pythagorean tuning scheme and then
adjusted it to make it more palatable.

Successive instructions on fretting from the mid-sixteenth century onward de-emphasized
Pythagorean systems in favor of tempered intervals and methods of equal fretting.
While they certainly acknowledged Pythagorean intervals, and writers such as Bermudo
and Ganassi offered complete fretting systems using Pythagorean ratios, they were never
intended to be the final solution. Most musicians recognized the importance of
Pythagorean intervals, such as the fourth, fifth, octave and in some cases the
wholetone; however, they also used non-Pythagorean intervals to create a better
solution.  The primary distinction between sources that advocated Pythagorean tuning
over a tempered kind was that Pythagorean sources tended to focus more on
mathematical issues than musical ones.


%
% Hans Gerle
%
\section{Gerle's fretting instructions}

The first major break with Pythagorean systems of tuning for lute came from Hans Gerle, a
lutenist and composer who published his treatise in 1533.  In addition to its specific
explanation of fret placement, it is also one of the best sources for lute instruction and
practice in the early sixteenth century.  It contains detailed explanations of tuning,
right-hand and left-hand technique as well as tablature.  The most important feature of
Gerle's treatment of lute fretting, and the others that I will be analyzing in this
chapter, is that Gerle was a lute player and not a theorist. From this viewpoint, matters
of theory are discarded in favor of practicality.  Gerle probably knew very little of the
implications of breaking with the tradition of Pythagorean tuning.  His emphasis was on
finding a practical way to fret the lute in such a way that it would sound agreeable.  The
importance of his instructions become more apparent when we see other lutenists such as
John Dowland borrowing many of Gerle's methods over 75 years later.

Gerle's instructions are directed towards a player who may or may not have theoretical
knowledge of music, but would have had some rudimentary knowledge of arithmetic and
geometry.  His process involved using a compass and a straight edge, such as a piece of
wood or other material that the player cuts to be the same length as the vibrating string
length of the instrument.  Marks were made on the straight edge, using the compass to
divide the distances between each mark into different parts.  Once all the appropriate
marks were made, they could be transferred to the fretboard and the frets placed
accordingly.

He uses a very straight-forward, step-by-step approach for setting each fret, beginning
with the twelfth.
\begin{blocks}
Take a straight-edge that is thin or else a flat piece of wood like a ruler, and make it
of such a length that at the top it touches the piece of wood that the strings lie on and
also touches the bridge that the strings lie on, and when you have made the ruler so that
it touches at both ends (don't make it too short; it must touch as I have said), mark the
bottom part at the bridge with an a, and the top part with a b, so that you will know
which end belongs to the bridge.  Then lay the ruler on a table, and take a compass and
find the middle of the ruler.  Mark it with a point or little dot and put an m there.
\end{blocks}
The letter \textit{a} marks the bridge and the letter \textit{b} marks the nut.
By placing the twelfth fret in the middle of the string, he divides the string in half
yielding $ \frac{1}{2} $, or in terms of a ratio, this would be an octave of 2:1.
Compasses were widely used at this time to perform all kinds of geometrical divisions and
we can assume most learned individuals at this time would know how to use one in order to
execute Gerle's instructions.  Euclid, the Greek mathematician, described procedures for
dividing lines into equal parts in his famous work \textit{Elements}.  This book was
first translated into Latin in 1482 and appeared in English translation by 1570.  Although
Gerle was German, it is reasonable to assume that he could have had access to either a
copy in Latin or one in German.  As far as we know, Gerle was not educated at a university
and probably had little or no knowledge of Latin; however, it is still reasonable to
assume that with Euclid's work in circulation, he could have come by the knowledge needed
to perform the required geometrical calculations.

% Gerle's seventh fret
Gerle continues using pure intervals for the fifth, or seventh fret:
\begin{blocks}
Then divide from the m to the b [in] three parts; and the first part from the m gives you
the seventh and lowest fret.  Mark it with a dot and put the number 7 there.
\end{blocks}
The letter \textit{b}, marked in the previous step, denotes the nut at one end of the ruler.
Gerle now switches to numbers for frets and indicates the seventh fret with the number 7.
The fret is marked at the first of the three parts starting from the twelfth fret or
middle of the string.  The resulting fret placement is one-third the length of the string,
making the vibrating length of the remaining string two-thirds. We can express this
semantically as a string divided into three parts with a vibrating length of two parts, or
3:2, which corresponds to the pure fifth.  Gerle could have divided the entire string into
three parts and placed the seventh fret at the first part from the nut, but it seemed more
important to base successive frets on the location of existing ones, in this case, the
twelfth.

% Gerle's first fret
Gerle's instructions continue with the first fret, and similar to his instructions for the
seventh, he builds on the calculations of the previous fret to find its location.
\begin{blocks}
Then divide elevenfold from the number to the b, and two of the same parts down from the b
give you the first fret.  Mark this also with a dot and put the number 1 there.
\end{blocks}
The ``number'' he is referring to is seven, or the seventh fret that we just marked in the
previous step.  Here he has us divide the distance from the nut to this fret in eleven
parts and mark the second of these parts starting from the nut. This creates a ratio of
33:31, which is not Pythagorean but instead somewhere between the diatonic semitone
16:15 and the ratio 17:16. Mark Lindley has determined that this distance is actually a
one-sixth comma semitone, but what kind?  In chapter two, we examined systems of equal
division that divided the octave into multiple parts greater than twelve.  Sixth-comma
meantone corresponded to an octave divided into 55 parts, and the first fret on the lute
would be the first semitone in a 55-division system.  If the semitone is chromatic, 
it has four parts, and if it is diatonic, it has five.

To determine which semitone Gerle was using we can compare our calculations of his fret
distance with the known values of different semitones.  If we express Gerle's 33:31 ratio
as a decimal by dividing 33 by 31, we get $ \frac{33}{31} = 1.0645 $ rounded off
to the nearest fourth decimal place.
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r l }
        Equal temperament semitone:     & $ 2^\frac{1}{12} = 1.0600 $ \\
        Sixth comma chromatic semitone: & $ 2^\frac{4}{55} = 1.0517 $ \\
        Sixth comma diatonic semitone:  & $ 2^\frac{5}{55} = 1.0650 $ \\
    \end{tabular}
    \end{center}
    \caption{Equal and sixth-comma semitones}
    \label{table:6semitones}
\end{table}
Comparing that value to the values of other semitones in table ~\ref{table:6semitones}
we can see that Gerle's first fret matches the diatonic semitone in sixth comma
meantone temperament.

Gerle has departed from Pythagorean tuning and opted for a tempered semitone at the first
fret.  This is not to say that he is advocating a sixth comma meantone tuning overall.
Gerle still uses a pure fifth at the seventh fret, which is not the same as a
fifth tuned in sixth comma.  A sixth-comma fifth is slightly flatter.
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r l l }
        Pure fifth:        & 3:2                 & $ = 1.5000 $ \\
        Sixth comma fifth: & $ 2^\frac{32}{55} $ & $ = 1.4967 $ \\
    \end{tabular}
    \end{center}
    \caption{Comparison of pure and sixth-comma fifths}
    \label{table:6fifths}
\end{table}
The fifth is comprised of three whole tones and one diatonic semitone. In a sixth-comma
system this makes up 32 parts or \textit{dieses}, and the resulting interval is slightly smaller
than the pure Pythagorean fifth, which is what we should expect in any regular meantone
temperament (see table ~\ref{table:6fifths}). This will have implications when we examine
the internal tuning issues of lutes later.

% Gerle's second fret
After having placed the first fret, Gerle continues with the placement of the second, or the whole tone.
Here, he returns to the standard Pythagorean 9:8 ratio:
\begin{blocks}
Then divide again from the number 7 to the b threefold, and the first part down
from the b gives you the second fret.  Mark it also with a dot and put the
number 2 there.
\end{blocks}
Here we have a three part division from the seventh fret to the letter \textit{b}, 
which was marked in the first step at the nut.
By returning to our earlier calculation of the seventh fret, we can formulate the
second fret distance and vibrating length accordingly which results in the Pythagorean
wholetone.

% Gerle's fifth fret
So far, Gerle has mixed Pythagorean pure intervals with tempered ones, such as
the fifth, the major second and the tempered minor second.  As he moves to the
perfect fourth at the fifth fret, he writes:
\begin{blocks}
Then divide from the m to the b in two parts, and the one part gives you the
fifth fret. Mark it with a dot and put the number 5 there.
\end{blocks}
Starting from the twelfth fret, Gerle divides the distance into two parts and takes the
first part from the nut as the mark for the fret. The letter m is our octave fret, the
exact midpoint of the string. Since he is now dividing that again in half, this
produces a Pythagorean interval of the fourth with a ratio of 4:3.

% Gerle's sixth fret
Now Gerle turns his attention to the tritone, or sixth fret:
\begin{blocks}
Then put the sixth fret in the middle of the fifth and seventh frets.  Make it
with a dot and put the number 6 there.
\end{blocks}
If we assume that Gerle is using ``middle'' to denote the arithmetic mean, as opposed
to the geometric mean, we can calculate the middle distance between the
fifth and seventh frets and get a vibrating ratio of 24:17.  However, an alternative
explanation is that Gerle is being intentionally vague here as if to tell the player to
put the fret somewhere between the two frets, and not necessarily exactly in the
middle.  If we assume this, then it seems reasonable to assume that the player
could also alter the placement until it produced a good result.

Regardless of our interpretation of Gerle's instruction, the tritone is neither
Pythagorean ratio nor sixth-comma meantone, but it may be close enough to be considered
equally tempered. In a sixth comma meantone with a 55-division octave, the tritone
contains either 27 or 28 dieses depending on whether the semitone above the
fourth is chromatic or diatonic.  For a lute tuned in G, this would be the difference
between a D$\sharp$ and an E$\flat$. Calculating the decimal equivalents for these two
semitones shows us that Gerle's tritone falls somewhere between the chromatic and
diatonic semitones.  Although this fret is not identical numerically to the
equally-tempered tritone, if we look at a scaled drawing depicting the location of the
fret in relation to the other intervals, Gerle's fret and the location of the
equally-tempered fret look nearly identical. We must note that Gerle places the fret in
the middle of the other two frets without the use of a compass.  Furthermore, while
Gerle's sixth fret may come close enough to equal temperament that we could call it an
equally tempered interval, he is not using equal temperament in the modern sense.  In
fact, Gerle was using meantone in its original sense, the way in which Aaron would have
tuned.

Meantone temperaments that did not use equal division had to approximate some of their
semitones by dividing the wholetone differently. For example, Aaron's
original instructions did not consider every semitone in the scale, it only covered a
set of thirds that were tuned pure leaving some of the other pitches indeterminate.
Semitones that were not included in the scheme were divided, splitting a wholetone into
two different parts. Since this was probably done by ear and without the aid of exact
measurements, the calculation of this fret is an approximation and we should discount
the use of arithmetic to determine the quality of the semitone between the fifth and 
seventh frets.

In order to better visualize frets that are approximations instead of
exact calculations, we can can look at a scaled drawing comparing the placement of
the different semitones.  I will be using many such scaled drawings and each will use
the same sample mensur length of 70 centimeters.  The purpose of the drawing is not to
show the exact placement of frets that would normally be approximate, but to
demonstrate how one type of inexact fret may be similar to another type of fret that is
exact in its placement by showing how close they are to another.  The choice of mensur
length is based on what best appears on the page.  While 70 centimeters is unusual for
a lute, a more common mensur length of 60 centimeters is smaller on the page and does
not exhibit the differences of semitone size as clearly.

In turning to Gerle's sixth fret, we can look at one of these scaled drawings comparing
his fret with the known placement of the chromatic sixth-comma semitone, the diatonic
sixth-comma semitone and the equally tempered semitone.  These frets are shown in red
while Gerle's frets are in black.
\input{figures/gerle-6}
Gerle's fret favors the chromatic side, although given the proximity of their
locations, his fret is very near the equally tempered semitone that it could be
considered closer. Another point to bear in mind is that as the mensur length
decreases, so will the distances between these frets.  Conversely, the distances will
increase as the mensur increases.  This will have more bearing on the theorbo, which
has a mensur length of 80 centimeters or more, and is discussed in later chapters where
I focus on executing different types of fret placements.

Gerle next instructs us where to place the third fret by taking a portion of the 
distance from the nut to the first fret.
\begin{blocks}
Then divide from the number 1 to the b [in] three parts, and when you have the
three parts then go with the compass unaltered down from the number 1 again five
spans; that gives you the third fret.  Mark it with a dot and put the number 3
there.
\end{blocks}
Gerle's uses the term ``span'' to denote one of the three parts in the distance from
the first fret to the net. Therefore, the total spans from the nut (b) to the third fret 
is eight: the five spans
from the first fret to the third, plus the original three spans from the nut to
the first fret.  The resulting 99:83 ratio is closest to a diatonic sixth comma
meantone minor third.
\input{figures/gerle-3}
In looking at the scale drawing, the black line representing the fret completely
occludes the red line of the diatonic sixth-comma semitone, indicating that for all
practical purposes the two semitones are the same.  Longer mensur lengths might exhibit
a slight difference in distance but it would be minimal.

The last fret is the fourth which makes the major third.  Similar to the sixth fret, he
averages the distance between the third and fifth.
\begin{blocks}
Then put the fourth fret between the third and the fifth frets.  Mark it with a
dot and put the number 4 there.
\end{blocks}
Averaging the distance between the two frets results in the ratio 792:629, which comes
closest to the equal tempered major third. Referring to a drawing of the fourth fret,
Gerle's major third is flatter than a true equally tempered third, but sharper than a
major third in a sixth comma temperament.
\input{figures/gerle-4}
With our sample mensur length of 70 centimeters, the actual difference
between Gerle's major third and a sixth comma meantone major third is about 2
millimeters.  The difference between Gerle's third and a quarter comma meantone third
is 4 millimeters. So when speaking of comparisons that account for a few thousandths of
a decimal, the numbers themselves are small, but when translated into a real
instrument, the differences become large.

Gerle limits his fret placement calculations to the seventh fret only; however,
he does provide instructions for placing an eighth but without leaving us any
geometrical calculations to do so.
\begin{blocks}
But if on the lute one wants eight frets, then let him make the eighth fret a
little closer to the seventh fret than the sixth is.
\end{blocks}
Taken at face value, we cannot really determine what Gerle exactly means other
than the semitone between seventh and eighth fret is smaller than the semitone
between the sixth and seventh.  At this point, we are to assume that the player
may use his or her ear to adjust the fret accordingly.  This same rule would
hold true for the rest of the frets as well.

Taken as a whole, Gerle's fretting instructions are a mixed assortment of different
intervals.  He uses Pythagorean ratios for the basic intervals, but relies on tempered
ones for the others.  These are either sixth-comma meantone in nature, or completely
unique.  His approach seems to have taken hold with other players because we see the
same kind of heterogeneous technique used for many years afterwards.

%
% John Dowland
%
\section{John Dowland's fretting instructions}

Gerle's fretting instructions seemed to have survived for quite some time after his
death because they appear in an almost identical form an an instruction method
by the English lutenist John Dowland. One of the best known lutenists of his day,
Dowland was an important composer of songs and solo music for the lute and remains so
to this day. During his lifetime, he published five books for lute and voice, as well
as music for viol consort and lute accompaniment.  His solo works existed mostly in
manuscript, with the notable exception of the \textit{Varietie of Lute-Lessons} which was
published by his son, Robert Dowland, in 1610 and contained music by John Dowland and
some of his contemporaries.

As the title suggests, the \textit{Varietie} was intended as an instruction book, however the music
in it demands a high level of ability.  The book begins with two prefaces, one written by John
Baptiste Besard and a second by the elder Dowland. Besard's preface, ``Necessarie Observations
Belonging to the Lute and Lute-playing,'' was published three times elsewhere: First, in Latin, in
Besard's own \textit{Thesaurus Harmonicus}, a compendium of lute music published in 1603, 
where it was titled ``De modo in testudine studendi libellus.''  The second time was in 1617,
where it was revised slightly and published in his \textit{Novus Partus} with the title ``Ad artem Testudinis brevi.''
Lastly, it was published again that same year, in German, in a pamphlet entitled \textit{Isagoge in artem institutionem}.
Each version contained minor differences, but focused on technique instead of tuning or temperament.

Dowland's own instructions, entitled ``Other Necessary Observations belonging to the
Lute'' offered additional advice on choosing lute strings and a method for setting the frets on the
instrument that followed Gerle's very closely. While his instructions postdate Gerle by more than 70
years, it is obvious that Dowland must have known of Gerle's book because of the similarity of their
instructions.  It is also possible that they consulted a similar source on geometry and measurement
and reached the same conclusions, but it is more likely that Dowland simply knew about Gerle's
instructions either through his book directly or from common practice and was repeating them here.

After giving the usual historical account of Pythagoras discovering harmony by
listening to the hammers of blacksmiths, Dowland instructs us to procure a piece
of ``whitish'' wood that is just as long as the distance from the inward
side of the nut to the inward side of the bridge on the lute.  From there, his
fretting instructions are exactly the same as Gerle's except that
Dowland specifies letters instead of numbers since he was using the French style
tablature system.

If we account for several printing errors that appear in Dowland's instructions, they
are very straightforward.
\begin{blocks}
Wherefore take a thinne flat ruler of whitish woode, and make it just as long and
straight as from the inward side of the Nut to the inward side of the Bridge, then note
that wnd [\textit{sic}] which you meane to the Bridge with some small marke, and the other end with
the letter A, because you may know which belongeth to the one and to the other. then
lay the ruler upon a Table, and take a payre of compasses and seeke out the just middle
of the Ruler: that note with a pricke, and set the letter N. upon it, which is a
Diapason from the A. as appeareth by the striking of the string open.
\end{blocks}
After describing the same manner of using a ruler in place of the mensur length of the
string, Dowland places the twelfth fret, fret N, at the midpoint of the string creating
a pure 2:1 octave.
\begin{blocks}
Secondly, part the distances from N. to D. in three parts, then the first part gives
you the seaventh fret from the Nut, making a Diapente: in that place also set a pricke,
and upon it the letter H.
\end{blocks}
Dowland makes the first of many typographical errors and uses the letter D instead of
A. If we allow for that mistake and read the passage 'N. to A.' and not 'N. to D.', it
is obvious that Dowland is marking a perfect 3:2 fifth at the letter H.  When he sets
the first and second frets, he repeats Gerle's measurements almost verbatim, yielding a
33:31 diatonic sixth-comma semitone and a 9:8 Pythagorean wholetone.
\begin{blocks}
Thirdly, devide [\textit{sic}] the distance from the letter H. to the letter A. in eleaven parts: two
of which parts from A. gives the first fret, note that with a pricke, and set the
letter B. thereon, which maketh a Semitone.  Fourthly, divide the distance from H. to
the letter A. in three parts, one of which parts from A. upward sheweth the second
fret, note that with a pricke, and set the letter D. upon it, which maketh a whole Tone
from A.
\end{blocks}
When he gets to the fifth fret there is another error, mistakenly indicating the first
fret instead of the fifth, but otherwise the same 4:3 pure fourth that Gerle uses.
\begin{blocks}
Fifthly, divide the distance from N. to A. into two parts, there the first part sheweth
you the first [ie. fifth and not first] fret, sounding a Diastessaron: in that place
also set a pricke, and upon it the letter F.
\end{blocks}
The sixth fret is placed the same as Gerle's, as the mean distance between the fifth and
seventh, giving us the same ratio of 24:17.
\begin{blocks}
The sixt fret with is a G. must be placed just in the middle betwixt F. and H. which
maketh a Semidiapente.
\end{blocks}

When Dowland reaches the third fret, he departs from Gerle's geometrical instructions:
\begin{blocks}
Seventhly, divide the distance from the letter B. to A. in three parts, which being
done, measure from the B. upwards foure times and a halfe, and that wil give you the
third fret, sounding a Semiditone: mark that also with a prick, \& set thereon the
letter D.
\end{blocks}
Whereas Gerle would measure five of his spans from the first fret, Dowland only
uses four and a half. This results in a ratio of 198:168 which in our sample diagram
appears the same as the sixth comma chromatic semitone.
\input{figures/dowland-3}
Dowland is intentionally vague here.  Instead of using whole numbers as everyone else
does when making geometrical divisions with a compass, he uses fractions. If we wanted
to be exact as, we could find the mean of the original span that is used to
find the distance to the third fret and have a true ``four and a half spans,'' but
Dowland does not instruct us to do that.

There are several possible explanations for Dowland's odd calculation:  First, Dowland
could have been leaving the method of calculation to the player.  Someone able
to execute Dowland's geometrical calculations could have found the
exact number of spans necessary, if he or she wished.  Perhaps
Dowland simply did not want give the extra detail. A second explanation is that
Dowland knew he preferred a chromatic semitone and used the specification of 4.5 spans
as an approximation.  The compelling case for this explanation is that the 4.5 span
measurement is so close to a chromatic semitone in sixth comma meantone, the same kind
of regular meantone temperament that Gerle was using. A third explanation is that
Dowland did not know he was expressing the difference between the chromatic and
diatonic semitone and arrived at the interval ``by ear.''  Either by experimentation or
estimation, he found a way to represent the fret placement mathematically with an
approximate measure of 4.5 spans.

% Dowland's fourth fret
As Dowland moves on the fourth fret, he places it using the same arithmetic mean that
Gerle used in his instructions.
\begin{blocks}
then set the fourth fret just in the middle, the which wil[l] be a perfect
\emph{ditone}:
\end{blocks}
Since Dowland's third fret was slightly smaller than Gerle's, this will make the
placement of his fourth fret different from Gerle's as well, coming to a ratio of
1584:1266, which is very close to a quarter-comma chromatic semitone.
\input{figures/dowland-4}
Although this could be an argument in support of quarter-comma meantone on
fretted instruments, it is the only interval in this scale that appears to be tempered
this way.  The other intervals are still a mix of Pythagorean and sixth-comma meantone
intervals.

% Dowland's other frets
Dowland goes on to give us instructions for placing the eighth, ninth and tenth frets
instead of stopping at the seventh as Gerle does; however, it is at this point that his
calculations go awry.  He places the remaining three frets using the same process:
\begin{blocks}
then take one third part from B. to the Bridge, and that third part from B. maketh I.
which soundeth \emph{Semitonium cum Diapente}, then take a third part from the Bridge
to C. [N.B. He means C. to the Bridge and not the other way around] and that third part
maketh E. [N.B. Another misprint here, he means the ninth fret, K.] which soundeth
\emph{Tonus cum diapente}, or an \emph{Hexachordio maior}.  Then take one third part
from D. to the Bridge, and that third part from D. maketh L.  which soundeth
\emph{Ditonus cum Diapente}.
\end{blocks}
Apart from the errors with the ninth fret, Dowland is repeating the same pattern for
each remaining fret.  He seems to be calculating their placement by dividing the
total distance between one fret and the bridge into three parts and then setting the
fret at one-third of the distance.  In looking at the mathematical result of
this process, we find the ratios of the higher frets are actually decreasing when they
should be increasing.  Either we are misinterpreting Dowland's instructions because of
further errors in the printed source, or Dowland has simply got his calculations wrong.

\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r l l }
        99:68 ratio   & $ \frac{99}{68}   $ &  = 1.4559 \\
        27:19 ratio   & $ \frac{27}{19}   $ &  = 1.4210 \\
        594:426 ratio & $ \frac{594}{426} $ &  = 1.3944 \\
    \end{tabular}
    \end{center}
    \caption{Decimal ratios of Dowland's eighth, ninth and tenth frets}
\end{table}

If we return to the calculations that we obtained following Gerle's scheme, the only
differences between those and Dowland's are the third and fourth fret. However, in
looking at the semantic instructions themselves the real difference lies with the third
fret.  Gerle required 5 spans from the first to the third fret while Dowland required
slightly less, 4.5.  The semantic instructions for the fourth fret were the same in
both sources: place the fourth fret in the middle, between the other two.  The ratio of
the fourth fret was only different because the third fret had been shorted.  
If we then entertain the possibility that Dowland was essentially using
Gerle's original instructions directly, but with only a slight modification to the
third fret, perhaps due to personal preference, we could then see that the eighth,
ninth and tenth frets were of Dowland's own doing.

The last three frets of Dowland's scheme had no precedent, if we assume that Dowland
was using Gerle as his primary source.  Dowland then was on his own in determining in a
proper procedure for obtaining the right proportional distance for those last frets.
Due to the calculations above, it is clear that one cannot arrive at a usable
placement for the last three frets of his scheme. Nor can one use Dowland's own
fretting instructions as complete source because there is no mention of the eleventh
and twelfth frets which figure prominently in his music, and especially the contents of
the book which his instructions preface.

\section{Ganassi's \textit{Regola Rubertina}}

Because Dowland and Gerle hold many similarities in their fretting instructions, we can
make a strong case for using them as a single unit for comparison, since all but one of
their frets align very closely.  An important source that would contrast with their
methods is Silvestro Ganassi's instruction method for viola da gamba entitled,
\textit{Regola Rubertina}. Although he was a gamba player and his method is intended
for that instrument and not lute, it does contain fretting instructions that are
valuable when applied to the lute

Ganassi approches his subject slightly differently than Gerle and Dowland.  In the
fourth chapter of his treatise, he specifies the locations of each fret using a compass
and string division, just as Gerle and Dowland do; however, in the fifth chapter, he
adjusts some of these frets to different positions by comparing unisons from one string
on the instrument to another.  The first, second, third, sixth and eight frets are
adjusted flatter than their initial placements, while the fourth fret is adjusted
sharper.  One of these changes is markedly different from other treatises of the
time, namely the adjustment of the fourth fret; although, the frets that make the wholetone, fourth and
fifth, are all kept to their Pythagorean ratios.

For example, Ganassi opens his fourth chapter with the second fret, and uses the same
Pythagorean wholetone ratio that everyone else has used at this time:
\begin{blocks}
Please note that the proportion \emph{sesquioctava} produces pitches expressed
by these two number, 9:8.  This proportion determines the location of the second
fret.  If you divide the string, beginning from the nut on the fingerboard and
ending at the bridge, where the bow is drawn, into nine parts, the first of the
nine parts sets the boundary of the second fret.
\end{blocks}
He also uses the pure 4:3 ratio of the
perfect fourth because he says the open strings that are played against it must
be in unison.
\begin{blocks}
Then, you divide the string into four parts; the first of these four part will
set the location fo the fifth fret, which produces the consonance of a fourth,
which is created by the proportion of sesquitertia indicated by a ratio of 4:3
[ \ldots ] This produces the consonance of the perfect fourth, because if
one then plays the open string, which is at the end of the fourth part of the
string length, one achieves the opposite of the 4:3 ratio
\end{blocks}
Most players would have tuned their open strings using the the fifth fret of the
adjacent string.  In this case, Ganassi is merely underscoring the fact that the fifth
fret should be at the ratio of a pure fourth because the interval of a fourth between
the adjacent strings of the viol should be pure. The seventh fret is kept to a pure
ratio as well:
\begin{blocks}
Then you divide the string length into three parts. The first of the three parts will
be the end of the seventh fret, thereby producing the consonance of the perfect fifth,
of \textit{diapente}, which is formed with the proportion of \textit{sesquialtera}
indicated by the ratio of 3:2.
\end{blocks}

So for these frets, Ganassi is reiterating what most other theorists believed about
tuning the wholetone, fourth and fifth by maintaining their Pythagorean ratios. In
chapter five, ``Method of Adjusting the Frets,'' Ganassi verifies the positions of
these three frets by checking the unisons against the various open strings of the
instrument.  All this is to ensure that he wants the all the intervals between open
strings to be pure Pythagorean ratios.  It is important to note here that this
precludes the use of any meantone temperament since the fifths would need to be
tempered flatter than pure.  We may also say the same thing about Dowland and Gerle's
methods since their fourths and fifths are pure as well.

Once Ganassi has completed setting those frets, he does not revisit them.  However,
this is not the case for the other frets of the instrument.  It is the other intervals
that distinguish Ganassi's ideas about fret placement from some his contemporaries.
Namely, he departs from certain meantone and Pythagorean ratios in favor of his own
that are produced by comparing tones from one string and fret to the tones of a
different string or fret.

\subsection{Ganassi's non-Pythagorean frets}

Ganassi's first fret differs in size from both Gerle and Dowland.  He begins by having
us place the first fret in a Pythagorean ratio, but eventually moves it to a different
kind of ratio.  In chapter four, he writes:
\begin{blocks}
After you have positioned the second fret in the manner described above, the
first fret should be set half way between the major and minor semi-tones by
their respective proportions.  In order not to go into this at length, however,
I believe that I have chosen a similar method for finding the first fret which
produces a minor semitone, quite easily.\autocite[106]{RB:2}
\end{blocks}
Another translation of the same passage from Mark Lindley's books is slightly different:
\begin{blocks}
\begin{center}
\begin{tabular}{r l}
Dapoi che hauerai trouato \&           & Now when you have found and \\
terminato il ditto sec\={o}do tasto    & located the second fret \\
c\={o} il modo ditto di sopra          & by the method given above, \\
il tasto primo sera terminato al meza  & the first fret is located halfway  \\
tra il scagneleto del manico al        & between the nut and the  \\
secondo tasto ma de piu zoe            & second fret, but more, i.e.  \\
batt\={e}do di fora la mita della      & down the neck by half the \\
grosseza del tasto . . . \& in questo  & fret's width.  In this regard \\
ti haueria possuto resonar il          & I could have calculated for you the \\
partimento del semiton maior al minor  & division of the major and minor semitones. \\
. . . il primo tasto elqual fa leffeto & The first fret gives the effect \\
del semiton minor . . .                & of the minor semitone. \\
\end{tabular}
\end{center}
\end{blocks}

Taking both translations into account, the placement of the first fret appears to be
halfway between the \textit{scagneleto del manico} or nut and the second fret.  When
the Pythagorean wholetone (9:8) is divided in half, the resulting ratio is 18:17, or
what is the closest to an equal semitone.
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r l l }
        18:17 semitone:             & $ \frac{18}{17}  $ & = 1.0588 \\
        Equal temperament semitone: & $ 2^\frac{1}{12} $ & = 1.0595 \\
    \end{tabular}
    \end{center}
    \caption{Comparison of the 18:17 and equally tempered semitones}
\end{table}
It is not exactly the same as a true equally tempered semitone, but is close enough
that many theorists and players advocated using a system of semitones based solely on a
series of 18:17 divisions.

In the sixth chapter, Ganassi changes the position of the first fret yet again,
using the fifth fret of the fourth course as the reference:
\begin{blocks}
Having tuned the open fourth string and having adjusted the fourth fret, you should
then check the fifth fret of the third string. Its pitch will provide the means of
regulating the first fret of the fourth string. \autocite[114]{RB:2}
\end{blocks}
Ganassi numbers the courses of his instrument in the reverse order that we do
today: his ``first string'' is our sixth course, and his ``sixth string'' is our
first course. On a lute in Renaissance G tuning, the  above description would
correspond to the unision B$\flat$ found between the fifth fret of the fourth
course and the first fret of the third course. Because this unison will not be
in tune if the first fret is a 18:17 ratio, the fret must be moved towards the
nut, rendering it slightly flat. We can determine the  decimal equivalent of
this new adjusted first fret by referring to a chart of frets and their
placements from Richard Bodig's article on Ganassi's \textit{Regola}.
\autocite[67]{RB:3} He calculates new values for each fret using Ganassi's
instructions on where the frets are moved after they are initially placed. The
resulting decimal values occur once the player moves the frets to where they are
exactly in unison with one another according to Ganassi's instructions.

Using Bodig's calculations of Ganassi's fret adjustments, and comparing them to the
fret placements we have seen thus far with Gerle and Dowland, Ganassi's first fret is
much flatter than either Dowland or Gerle.  Curiously, Ganassi's adjusted first fret is
almost exactly the same as the chromatic sixth-comma semitone.
\input{figures/ganassi-1}
This differs from Dowland and Gerle whose first frets come to match the diatonic
sixth-comma semitone.  For a lute tuned in G, this would be the difference between a
G$\sharp$ and an A$\flat$ on the first string. While Ganassi's adjustment renders
the first fret in unison with the fourth, it is not a practical solution for a lute
because the pitches at the first fret are now G$\sharp$ for the first course, as well
as A$\sharp$ for the third.

Ganassi's third fret undergoes a similar transformation.  In the fourth chapter, he
sets the fret at a pure minor third.  Later in the sixth chapter, he adjusts this
and tunes it to the octave formed between the first fret of the third course and the
third fret of the sixth course. Returning to Bodig's calculations, this results in
lowering Ganassi's third fret from its initial pure minor third to something that is
flatter than an equally tempered minor third.
\input{figures/ganassi-3}
The drawing for this fret shows that it is approaching the chromatic sixth-comma semitone
of Dowland, but not close enough to be considered equal to it, nor to any other kind of
semitone that we have seen thus far.

Ganassi's initial setting of the fourth fret is halfway between the third and
fifth frets, before the third is adjusted.  This gives us a very strange ratio of
48:38.  Later, he adjusts the position of the fret so the octave between the second
fret of the sixth string and the fourth fret of the fourth string are in tune.
\input{figures/ganassi-4}
As the drawing indicates, his fourth fret is almost unusable since it is yielding a
major third that is even sharper than equal temperament.  Ganassi provides no further
explanation for this, but we must assume that most players would change this to
something more suitable and dismiss his instructions for this particular fret.

The initial placement for the remaining sixth and eighth frets is similar to the
fourth, and like his contemporaries, Ganassi does not provide any information about
frets beyond the eighth.  His sixth is the same as Gerle's, the arithmetical mean of the
distance between the fifth and seventh, or a ratio of 24:17, and the eight fret, which
is the most specific we have seen in any source thus far, results in a 8:5 ratio.
\begin{blocks}
Then the sixth fret is set at the midpoint of the space between the fifth and seventh
frets but somewhat less, that is so that the thickness of the fret is within the
compass of the distance; that will set its position.  The eighth fret will be located
so as to have the same spacing as from the fifth to the sixth frets.
\end{blocks}
Ganassi as well as other sources at this time such as Dowland, mention the thickness
of the gut string used as the fret and allude to the fact that this would have
some impact on its placement.  For example, a particularly thick fret might offset
calculations by as much as a millimeter.  For modern lute players, this would be an
issue for the lower frets, where gut strings of a millimeter or more are often used, but
the thickness decreases to less than a millimeter for higher frets. No
information exists about the specific thicknesses of the gut strings used for frets on
instruments during this time.  Ganassi's mention of it here at least indicates that it
may have been an issue.  His instruction is simply to ensure that no matter how thick
the fret is, its entire thickness is \textit{within} the span of the compass.

The final adjustments to Ganassi's fretting scheme include the sixth and eighth frets,
which are moved slightly flatter than where they were originally placed during chapter
four. The unison between the sixth fret of the third string and the first fret of the
second is used to adjust the sixth fret as needed in order to make that sound in tune.  The
eighth fret is then adjusted so that the octave formed between the open fourth string
and the eighth fret of the third string is in tune.  Using Bodig's measurements, we can
see that the final placement of the sixth fret moves it just slightly past the mark for
the chromatic sixth-comma semitone, or C$\sharp$ on an instrument tuned to a relative
pitch of G.
\input{figures/ganassi-6}
It is hard to say if this was intentional or simply a coincidence of Ganassi's
adjustment procedures.  This is perhaps more evident in the placement of the
eighth fret, which is moved into a position that is not in any one particular
temperament, making it unique.
\input{figures/ganassi-8}

Even though Ganassi spends an entire chapter of this treatise on placing the frets
according to geometrical calculations, it is not until the sixth chapter that we see
the complete picture of his scheme.  Dividing the process as he does seems to
acknowledge the dichotomy between a theoretical system of string division and a
practical one that works in the realm of performance.  Beyond this Platonic
differentiation in tuning, we really cannot know what Ganassi's motivations are.

Ganassi holds many similarities with other sources.  The second, fifth and seventh
frets are the same as Gerle and Dowland.  His first fret, however, is noticeably
flatter than the Gerle/Dowland model, making it closer to a sixth-comma meantone
chromatic semitone instead of a diatonic one which Gerle and Dowland favor. Perhaps the
most important feature that Ganassi's fretting scheme shares with others from this
period is that it combines different types of semitones from different temperaments.  This
further indicates that lute temperaments were treated differently than temperaments from
other instruments.

\section{Spanish Vihuela Sources}

During the sixteenth century, there were a number of publications in Spain of music
and instruction on the vihuela.  The vihuela shared many similar characteristics with
the lute, but the most important to discuss is how its players grappled with the same
fretting issues that lutenists did. While the instruments were different, Spanish
vihuelists knew of the lute's repertory and technique.  For example, in Miguel
Fuenllana's \textit{Orphenica lyra}, published in 1554, he discusses the right-handed
``foreign style'' of playing, which refers to the thumb and index alternation of
playing practiced by lutenists at the time.

The schemes for setting frets on the vihuela come from two sources. The first, and
earliest of the two is Luis Milan's \textit{El Maestro}.  Milan was a well-known vihuela
composer and poet who seems to have spent most of his active life around Valencia.
Although his birth and death dates are not known, it is believed he died some time
after 1561 due to a reference in one of the composer's books of poetry, \textit{El
Cortesano}.\autocite[6]{LG:1}  Milan is best described as both a poet and bard who
would sing or recite his poetry to his own music.  He was active mostly at the court of
Valencia where he made his living working for the dukes of Calabria. \textit{El
Maestro} is his only musical publication but remains one of the most important vihuela
publications we have, not only for its musical content but for its information about
performance practice such as tempo and interpretive elements that are absent from
other contemporary treatises.

While \textit{El Maestro} contains only a few references to fret placement they are
very significant.  Milan's own remarks were not as exact as the others we have
seen.  He gave no measurements with which to mark frets on a straight edge, but he
did give some clues as to which frets he wanted adjusted to achieve a different
quality for certain notes.  In Gasser's monograph on performance practices in Milan's
music, he translates the following section from \textit{El Maeastro} where Milan
discusses changing the position of the fourth fret before the music for Fantasia 14.
\begin{blocks}
Whenever you play the fourth and third tones in those places through which the fantasia
moves, raise the fourth fret of the vihuela a little, so that the note of the fret
becomes strong and not weak.
[ Siempre que ta\~{n}erais el cuatro y tercero tono por estos t\`{e}rminos que est
fantas\`{i}a anda, alzar\`{e}is un poco el cuarto traste de la vihuela para que el
punto del dicho traste sea fuerte y no flaco ]\autocite[156]{LG:1}
\end{blocks}
Additionally, there is another piece of information concerning the same fret before the
beginning of the music for \textit{Con pavor record\'{o} el moro}.
\begin{blocks}
Playing in these pieces on the vihuela, you have to raise the fourth fret a bit toward
the pegs.
[ Ta\~{n}do por estas partes en la vihuela hab\'{e}is de alzar un poco el
cuarto traste hacia las clavijas de la vihuela ]\autocite[156]{LG:1}
\end{blocks}
Both references concern the fourth fret, and both offer an adjustment that raises the
fret closer to the pegs, making the interval smaller.  This would equate to a narrower
major third; however, it also assumes that the fret as been placed in some manner prior
to being moved.  It is not possible to know by what means Milan would have first placed
his fourth fret, but it seems that for certain pieces, he preferred it
to be slightly flatter than its initial placement.  This could corroborate Dowland's
use of a quarter-comma chromatic semitone for his fourth fret, which is substantially
flatter than either an equally-tempered or Pythagorean semitone.  However, the problem
here is that Dowland achieves a flatter fourth fret by also making his third fret flat
as well.  Milan mentions no other fret to adjust, so we might assume that all his frets
were placed according to standard Pythagorean proportions.  Still, both Dowland and
Milan do show a preference for flatter fourth frets.

Perhaps the most complete discussion on vihuela fretting appears in Juan Bermudo's
\textit{De ta\~{n}er vihuela}, published in 1555.  Bermudo's publication is one of the
most extensive on vihuela performance practice, including several chapters on
intabulation for the vihuela, and fret placement.  Bermudo provides a total of three
different fretting schemes.  The first of these, which appears in chapter 77, is
Pythagorean.  The second scheme attempts to correct the semitones in the first scheme
by adjusting the first, sixth and eighth frets while leaving the others in their
original Pythagorean form.  The third and last of Bermudo's schemes is unrelated to
either of the previous two and is the closest to modern-day equal temperament.

Dawn Espinosa translated Bermudo's work, with the original Spanish printed alongside
the English translation, and prefaced her translation with an informative discussion.
In it, she provides a table of Bermudo's fretting schemes and the ratios she has
calculated for each fret. For my analysis, I have used these to compare each of
Bermudo's fretting schemes with the others that we have seen so far.

Similar to Ganassi, Bermudo gives us a simple Pythagorean-based fretting system but
then offers two alternative systems as well. This may help explain the changes in
tuning occurring during the time.  Prior to the sixteenth century, Pythagorean tuning
had been the dominant system, but it had its inadequacies.  By starting with a
Pythagorean system and then providing alternatives, writers could now acknowledge its
importance and at the same time move past its insufficiencies.  For example, Bermudo's second
fretting system presents a solution that corrects some of the problems with
a Pythagorean system and arrives at the exact same placement for the third fret as
Ganassi does.
\input{figures/bermudo-3}
Also similar to many of his contemporaries, his solution for the first fret clearly
advocates the diatonic semitone.
\input{figures/bermudo-1}
However, it is slightly sharper than the sixth-comma diatonic semitone of Gerle and
Dowland.

Bermudo refers to the ``faults'' found in fretting systems between notes that are
\textit{mi} or \textit{fa}.  In terms of sixteenth-century theory, this was the
modern-day equivilent of a sharp note versus a flat note.  After setting the frets in
his first scheme, the Pythagorean one, he discusses the problems found on the eighth
fret:
\begin{blocks}
On the eighth fret there seem to be three faults.  This fret should be \text{fa} for the
seventh, fourth and first strings, but is it [made] \text{mi} for all the strings.
\autocite[95]{DE:1}
\end{blocks}
Bermudo is highlighting the central problem with fretting instruments: the distance of one
semitone on one string is the same for all the other semitones on that string.  Depending
on the temperament and semitone, a minor semitone might be required for one string while
a major semitone would be needed for another.

One of Bermudo's solutions, which we also see with Milan, is to adjust the frets
according to the key of the piece.  Milan, for example, would adjust the fourth fret
when playing the third and fourth tones, while Bermudo describes players who move their
frets according to their ears:
\begin{blocks}
We have seen players who, with their frets set for the sixth, want to play the fourth
mode, but are unable to do it without moving the frets as their good ear tells them.
What I intend to do here is to give the measurements by which to place the frets, so
that those who are not [such good] musicians will be able to place them with ease and
exactiude, and thus the vihuela with be more prefect.
\autocite[78]{DE:1}
\end{blocks}
Bermudo seems to suggest that it is the good players that are able to move their frets
for different modes, while inexperienced or lesser skilled players are unable and should
rely on Bermudo's calculations to place the frets for them.

For those players who are unable to place their frets correctly by ear, Bermudo's
ultimate solution is a scheme that matches modern equal temperament very closely.  His
third scheme essentially splits the wholetone into two equal halves, although the
composer himself acknowledges that it is not his aim to do so:
\begin{blocks}
All the [theorists] agree that the whole tone cannot be divided into two equal
semitones, but that is what we are presuming to do.  The above being the most agreed
and true thing among theorists, on the vihuela we find the opposite in practice.
\autocite[xx]{DE:1}
\end{blocks}
For Bermudo, the issue was a question of theory versus practice.  Espinosa summarizes
this issue succinctly: ``for him, theory has more authority than practice, but he
concedes that practice precedes theory.'' \autocite[xx]{DE:1}

In table~\ref{bermudo:differences}, the difference between each of Bermudo's fret and
modern equal temperament is expressed in both cents and in three different mensur
lengths: 600mm, 700mm and 800mm.  Millimeters are used for the mensur length because they
can express the differences with more accuracy.  700mm is the same as 70cm, which
has been the reference mensur length in our other diagrams. Cents, a logarithmic
measurement, are used as another point of comparison.  Cents are often used with
pitches in equal temperament because the system divides the octave into 1200 cents giving
each semitone exactly 100 cents, thereby expressing the equal nature of the tuning.
Subjectively speaking, the human ear can notice differences of a few cents, but near or
less than one cent would be difficult.
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r r| r r r }
      \textit{Fret} & \textit{In cents} & \multicolumn{3}{c}{\textit{At different mensur lengths in mm}} \\
       &  & \textit{600mm} & \textit{700mm} & \textit{800mm} \\
      \hline
              1  &          0.50	&          0.16	 &          0.19  &          0.22  \\
              2  &          0.31	&          0.09	 &          0.11  &          0.13  \\
      \textbf{3} & \textbf{-5.87}	& \textbf{-1.71} & \textbf{-2.00} & \textbf{-2.28} \\
              4  &          0.62	&          0.17	 &          0.20  &          0.23  \\
              5  &         -1.96	&         -0.51	 &         -0.59  &         -0.68  \\
              6  &         -1.46	&         -0.36	 &         -0.42  &         -0.48  \\
              7  &         -1.65	&         -0.38	 &         -0.44  &         -0.51  \\
      \textbf{8} & \textbf{-7.82}	& \textbf{-1.72} & \textbf{-2.00} & \textbf{-2.28} \\
              9  &         -1.33	&         -0.27	 &         -0.32  &         -0.37  \\
              10 &         -3.91	&         -.76	 &         -0.89  &         -1.02  \\
    \end{tabular}
    \end{center}
    \caption{Differences bewteen Bermudo's third scheme and true equal temperament}
    \label{bermudo:differences}
\end{table}
The negative numbers in the table show a fret that is flatter than equal temperament,
while a positive number indicates a fret sharper than equal temperament.

As the table shows, all but two of this frets are within a few millimeters of an
equally-tempered semitone, in a variety of different mensur lengths. The only
exceptions are at the third and eighth frets which are flatter than equal temperament.
When looking at the differences in cents, these frets are quite flat, and
most listeners would notice a difference of 5 to 8 cents; however, the differences of
fractions of one cent would be virtually undetectable.  For what reason Bermudo chose
to make these particular frets flatter than the others, by comparison, we cannot know.
Another interesting aspect to Bermudo's scheme is that, after the fourth fret, each
one is gradually flatter.  This may be to counteract a problem that we common with lutes
and fretted instruments where pitches at the higher frets tended to sound sharper. The
effect was greater at the highest frets, such as the tenth, which is perhaps why this
fret is almost four cents flat in Bermudo's scheme.

\section{Sources of Equal Fretting}

The last category of fretting systems that we will examine came from the late sixteenth
and early seventeenth centuries and advocated dividing the octave into equal semitones.
This is not to say that these systems produced the kind of equal temperament that we
have today. Creating a temperament with semitones that were exactly equal to one
another required a special class of mathematical functions called logarithms, which
were not used in tuning until later in the seventeenth century.  Despite their ability to
correctly divide the octave into equal semitones, equal temperament did not become a
true standard until the twentieth century.  As Ross Duffin has argued, even during the
nineteenth century musicians still tempered their fifths more than those in equal
temperament.

Prior to these advances in mathematics, equal semitones were approximated using other
methods. Sources during this time describe several different systems of measurement that
produced temperaments very close to modern-day equal temperament, but which were achieved
through a means of calculation that did not involve logarithms.  The first of
these we have already seen with Bermudo's third method of fretting, which closely approximates
equal temperament and uses standard geometrical string divisions common to
most lute sources at the time.  Additional sources on fretting in equal semitones come
from Vicenzo Galilei and Marin Mersenne, as well as numerous other anecdotal sources
collected by Mark Lindley in which contemporary musicians refer to the lute's
ability to tune equally.

One of the simplest ways in which a lute player could achieve equal semitones was to
divide the octave into a series of Pythagorean minor semitones, at a ratio of 18:17.
Called the \textit{18:17 Rule} because of this, both Mersenne and Galilei, as well as
numerous other sources, discuss this method.  The idea originates with the problem of
dividing the Pythagorean 9:8 wholetone into 2 equal parts, which I discussed in the
previous chapter.  A series of minor semitones produced fairly acceptable results, but
as the frets progressed further down the fingerboard, the distances became increasingly
smaller, making each fret flatter than the next.

Looking at table~\ref{18:17rule}, we can see the 18:17 rule compared with equal temperament
when executed with a variety of different mensur lengths as well as how it
compares in terms of cents.
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ r r| r r r }
      \textit{Fret} & \textit{In cents} & \multicolumn{3}{c}{\textit{At different mensur lengths in mm}} \\
       &  & \textit{600mm} & \textit{700mm} & \textit{800mm} \\
      \hline
      1  & -1.05  & -0.34 & -0.40 & -0.46 \\
      2  & -2.09  & -0.65 & -0.75 & -0.86 \\
      3  & -3.14  & -0.91 & -1.07 & -1.22 \\
      4  & -4.18  & -1.15 & -1.34 & -1.54 \\
      5  & -5.23  & -1.36 & -1.59 & -1.81 \\
      6  & -6.27  & -1.54 & -1.80 & -2.05 \\
      7  & -7.32  & -1.70 & -1.98 & -2.26 \\
      8  & -8.36  & -1.83 & -2.14 & -2.44 \\
      9  & -9.41  & -1.94 & -2.27 & -2.59 \\
      10 & -10.45 & -2.04 & -2.38 & -2.72 \\
      11 & -11.50 & -2.12 & -2.47 & -2.82 \\
      12 & -12.54 & -2.18 & -2.55 & -2.91 \\
    \end{tabular}
    \end{center}
    \caption{Comparison of fret placement using the 18:17 system}
    \label{18:17rule}
\end{table}
The differences between the two systems for the first five frets are very slight: only
less than two millimeters between the measurements of these frets at different mensur
lengths and at most a five cent difference in pitch.  At the higher frets, however, the
differences become larger.  By the time we reach the octave, the fret is more than
twelve cents flatter than the pure octave found in equal temperament.  This would be
very noticeable to most individuals.

Despite the decrease in distances, using a series of equal minor semitones in the 18:17
rule provided a practical temperament because its problems were mitigated by two
factors: First, most musicians at this time were using some form of temperament that
flattened fifths, and at the seventh fret where the fifth occurs, the fret is already
flattened by a few millimeters. Second, the tension created from stopping a string at a
given fret, especially if the fret itself is quite thick, or doubled as is the case
with the viola da gamba, can raise the pitch slightly. \autocite[21]{ML:1}  This effect
is compounded by the fact that changes in finger pressure or position have a more
noticeable effect on pitch at high positions. Making the frets gradually flatter as
they move to higher positions would help counteract this effect.

Another method of determining equal fret positions used a device called a mesolab,
which was an instrument of Greek origin that could determine mean proportionals, or the
average distance between two lines.  In 1558, Zarlino published a fretting scheme for
lute that relied on such a device. \autocite[26]{ML:1}  There is also a picture of one
in the \textit{Discours non plus melancoliques que diverses, de choses mesmement, qui
appartiennent a notre FRANCE: \& a la fin La maniere de bien \& iustement entoucher les
Lucs \& Guiternes}, published in 1556, which was mentioned earlier.  However, use of
the mesolab did not appear in any practical fretting guides and seems to be relegated
to music theory sources where equal temperament was a kind of puzzle to be solved and
not really taken seriously in a musical context.

By the seventeenth century, mathematics had advanced enough to employ decimal numbers
and logarithms to express musical ratios.  Johannes Faulhaber was the first to use
logarithms for a fretting scheme in his 1630 publication. \autocite[21]{ML:1}  In order
for it to work, the string had to be divided into 20,000 parts.  This idea later found
its way into an appendix to a translation of Rene Descartes' \textit{Musicae
compendium}.  Mark Lindley has concluded that its translator is William Brouncker, who
also wrote the appendix, which contains several fretting schemes for lute,
including our modern-day one using the $ \sqrt[12]{2} $ method, as well as others
of his own devising.

Despite the accuracy of logarithmic calculations in placing frets at equal temperament,
the method was largely ignored except in a few treatises.  While many authors spoke of
the lute's ability to play with equal semitones, the idea of an ``equal temperament''
was somewhat less certain.  From the available evidence, systems approximating equal
temperament and utilizing the idea of equal semitones certainly existed during this
period, but they still recognized the issue of somehow attempting to reconcile the idea
of unequal semitones found in the temperaments of all the other instruments.  The
central crux of the lute player's struggle was how to navigate between these two poles.

\section{Summary}
After this lengthy description of the available sources on fretting from the sixteenth
century and seventeenth centuries, some of them do agree on the placement of certain
frets (see table ~\ref{table:comparison}). The first, fifth and seventh are  almost all
uniformly Pythagorean in nature, making the fretted intervals of the perfect fourth and
fifth pure. When players tuned their open strings to one another by using the fifth
fret of the adjacent strings as a reference, this made the interval between the
two strings a perfect fourth as well.  This is corroborated by every source at this
time that contained instructions for tuning the lute's open strings.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ l l l l l }
      \textit{Fret} & \textit{Dowland} & \textit{Gerle} & \textit{Gnasi} & \textit{Bermudo II} \\
      \hline
      1 & \multicolumn{4}{|c|}{\cellcolor[gray]{0.9}diatonic sixth-comma} \\
      2 & \multicolumn{4}{|c|}{\cellcolor[gray]{0.9}Pythagorean 9:8 wholetone} \\
      3 & chromatic sixth-comma & diatonic sixth-comma & \multicolumn{2}{|c|}{\cellcolor[gray]{0.9}unique} \\
      4 & unique & unique & unique & equal \\
      5 & \multicolumn{4}{|c|}{\cellcolor[gray]{0.9}Pythagorean 3:2 fourth} \\
      6 & \multicolumn{2}{|c|}{\cellcolor[gray]{0.9}chromatic sixth-comma} & unique & unique \\
      7 & \multicolumn{4}{|c|}{\cellcolor[gray]{0.9}Pythagorean 2:3 fifth} \\
      8 & n/a & n/a & \multicolumn{2}{|c|}{\cellcolor[gray]{0.9}unique} \\
    \end{tabular}
  \end{center}
  \caption{Comparison of fretting schemes}
  \label{table:comparison}
\end{table}

The use of a Pythagorean wholetone for the second fret seems like a good choice at
first glance, since the second fret would be used in most scalar passages as the
wholetone above the open strings.  However, when used in chords, the second fret often
holds the third of the chord. In meantone temperaments, this means that the second
fret would need to be lower in pitch or moved closer to the nut in order to accommodate
either a sixth or quarter comma meantone temperament.  Furthermore, the use of pure
fourths between open strings, discounts the use of any meantone temperament because of
the tempered nature of their fifths.

With the remaining frets on their instruments, lutenists were experimenting with
meantone intervals.  These included frets one, three, four and six.
In the sources examined thus far, there is unanimous agreement that the first fret be a
diatonic sixth-comma semitone.  This would mean that for instruments tuned in either G
or A, their first frets would have to be A$\flat$ or G$\flat$ respectively. However,
there is disagreement over the remaining chromatic frets. For example, each
source has its own technique for placing the fourth fret.  Dowland and Gerle seem to
favor chromatic frets that closely approximate sixth-comma meantone intervals, while
Ganassi and Bermudo arrive at the same placement for their third fret, but neither of
them come close to any interval in the meantone scheme of temperament. Additionally,
the one glaring error that none of these sources seem to address is that tuning
open strings to perfect fourths would create a very sharp Pythagorean third in the
middle between the third and fourth courses.  It seems unlikely that anyone would have
tolerated that.

Given the available sources we have on fretting, we are unable to resolve all of the
discrepancies and we must inevitable make our own choices.  We can observe that players
embraced a variety of different techniques when addressing the issue and may use that as a
guiding principle when creating our own. Furthermore, their final solutions were not
uniform and no fretting system consisted entirely of Pythagorean intervals or meantone
intervals of one type or another.  The solutions were always mixed in nature.  With
that in mind, in the next chapter I shall propose different fretting
solutions based on the ones we have seen thus far, and examine them in a musical
context, using examples from the period.
